% ======================================================================
% RL algorithm definitions (single-step contextual-bandit framing)
% Requires: \usepackage{amsmath}
% ======================================================================

\paragraph{Setup.}
Each optimization step samples one parent state $s$, draws $G=64$ i.i.d.\ full
programs $a_i \sim \pi_\theta(\cdot \mid s)$, and assigns each a single terminal
scalar reward $R_i$. This is a one-step contextual bandit rather than a
multi-step MDP, so all three baselines reduce to score-function (REINFORCE-style)
estimators that differ only in how the advantage is formed. The action is the
entire token sequence, and the policy factorizes per token, so
\begin{equation}
  \nabla_\theta \log \pi_\theta(a \mid s)
  = \sum_{t} \nabla_\theta \log \pi_\theta\!\left(a_t \mid s, a_{<t}\right).
\end{equation}

\paragraph{REINFORCE (with a group baseline).}
The classic score-function policy gradient,
$\nabla_\theta J = \mathbb{E}\!\left[A \cdot \nabla_\theta \log \pi_\theta(a \mid s)\right]$,
using the group mean as a variance-reduction baseline and the within-group
$z$-score as the advantage,
\begin{equation}
  A_i = \frac{R_i - \operatorname{mean}_g(R)}{\operatorname{std}_g(R) + \varepsilon},
\end{equation}
which is exactly the GRPO advantage. There is no critic, no clipping, and a
single gradient step per batch. The estimator is mean-seeking: a rollout is
upweighted in proportion to how many standard deviations it lies above the group
average, so a single outlier-good construction receives only modest, normalized
reinforcement. It is unbiased given the baseline, since
$\mathbb{E}\!\left[b \cdot \nabla_\theta \log \pi_\theta\right] = 0$, but it has
the highest variance of the three.

\paragraph{A2C (advantage actor--critic).}
The same score-function gradient, but with a \emph{learned} state-value baseline
$V_\phi(s)$ in place of the batch statistic,
\begin{equation}
  A_i = R_i - V_\phi(s).
\end{equation}
Two coupled objectives are optimized: the actor ascends
$A_i \nabla_\theta \log \pi_\theta$, while the critic descends the regression loss
$\bigl(V_\phi(s) - R\bigr)^2$ toward the expected return. The general-MDP
advantage $r + \gamma V(s') - V(s)$ collapses here because there is no successor
state and no bootstrapping ($\gamma$ is irrelevant and every episode is one step
to termination), so $V_\phi(s)$ simply learns $\mathbb{E}[R \mid s]$ and the
advantage becomes $R - \mathbb{E}[R \mid s]$. The theoretical appeal over
REINFORCE is that a learned $V_\phi(s)$ generalizes across states and supplies a
baseline even for a state sampled only once. The asynchronous/synchronous
parallel-actor aspect of the A3C/A2C family maps onto the generation pool, but
the substantive difference from REINFORCE is purely learned-baseline versus
batch-baseline.

\paragraph{PPO (proximal policy optimization).}
The same advantage estimator as the REINFORCE control (the $z$-scored group
advantage), but with an off-policy correction and a trust-region bound so that
multiple gradient epochs can be taken over a single rollout buffer. Defining the
per-token probability ratio against the snapshot policy that generated the
rollouts, $r_t(\theta) = \pi_\theta(a_t) / \pi_{\theta_{\mathrm{old}}}(a_t)$, the
clipped surrogate is maximized:
\begin{equation}
  L(\theta) = \mathbb{E}_t\!\left[
    \min\!\Bigl(r_t(\theta)\, A,\;
    \operatorname{clip}\!\bigl(r_t(\theta),\, 1-\epsilon,\, 1+\epsilon\bigr)\, A
    \Bigr)
  \right].
\end{equation}
The clip is a first-order proxy for TRPO's KL trust region: it removes the
incentive to move the ratio beyond $1 \pm \epsilon$ on any token, so reusing the
batch for several epochs cannot push the policy far from the distribution that
produced the samples. The payoff is sample reuse with stability; the cost, on a
best-of metric, is that the clip deliberately throttles the per-step update.
Importance sampling also legitimizes the mild off-policyness already present:
the generation pool samples from a step-start snapshot shared across several
groups, so $r_t \neq 1$ even within a step, and the clipped ratio is the
principled correction.

\paragraph{Through-line.}
REINFORCE and PPO share a linear, mean-seeking advantage and differ only in
off-policy reuse and the trust region; A2C swaps the baseline source; and all
three are linear, whereas the entropic (TTT) objective is exponential and
max-seeking, which is the underlying reason it dominates on a metric that rewards
only the single best construction found.